plyranges: a grammar for transforming genomics data

### 2018-04-13 paper.Rmd
cohesion section draft

---
title: "plyranges: a grammar for transforming genomics data" 
author:
  - name: Stuart Lee
    email: stuart.lee1@monash.edu
    affiliation: Monash University
  - name: Michael Lawrence
    email: michael.lawrence@gene.com
    affiliation: Genentech
  - name: Di Cook
    email: di.cook@monash.edu
    affiliation: Monash University
address:
  - code: Monash University
    address: Department of Econometrics and Business Statistics, Clayton, Victoria, Australia
  - code: Genentech
    address: Bioinformatics and Computational Biology, Genentech, Inc., South San Francisco, California, United States of America 
abstract: |
    The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments. However, there is a cognitive load placed on a user in learning a data abstraction and understanding its appropriate use. Throughout a standard workflow, a user must navigate and know many of these abstractions to perform an genomic analysis task, when a single data abstraction, a _Ranges_ object will suffice. The _Ranges_ class naturally represent genomic intervals and their associated measurements. By recognising that the _Ranges_ class follows 'tidy' data principles we have created a grammar of genomic data transformation. The grammar defines verbs for performing actions on and between genomic interval data. It provides a principled way
    of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows. We have implemented this grammar as an Bioconductor/R package called plyranges.

@@

Another example of altering object semantics, while maintaining cohesion is
through the 'group_by' operator. Like anchoring, this operator decorates
a GRanges object with a column name (or names) that imply a partitioning of the 
GRanges by the unique values in the column(s). Functions defined in the 
`plyranges` DSL that perform restriction, aggregation, or column modfication
still perform those singular tasks, however grouping changes how those tasks
are performed.

A common operation to perform between two _Ranges_ objects is to find overlaps
or nearest neighbours. The _plyranges_ API recasts these operations as 'joins'
or 'pairing' operations. For overlaps, there are three join operations: 
`join_overlap_intersect()`, `join_overlap_inner()` and `join_overlap_left()`
which are shown in figure (\ref{fig:olap}).

@@

These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges objects
as metadata. The intersect join takes the intersect of the start and end coordinates
of overlapping intervals of the query and subject _Ranges_ 
(for _GenomicRanges_ it also accounts for sequence name), when there is a overlap
the metadata corresponding to the query and subject _Ranges_ are returned. 
Similarly, inner join takes the start and end coordinates of the query _Ranges_
that overlap the subject _Ranges_ and returns metadata of the overlapping query
and subject _Ranges_. Finally, the left join performs a left outer join between
the query and subject _Ranges_, it returns all genomic intervals from the query
ranges, and returns missing values in metadata columns when there is no overlap.

A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
`filter_by_non_overlaps()` and `group_by_overlaps()`. All overlap methods can
be modified with the `within` suffix (which changes the type of overlap 
from 'any' to 'within') or the `directed` suffix (which takes into account
the strand of a _GenomicRanges_ object.).

For nearest neighbours, the _plyranges_ API provides `join_nearest()`, 
`join_precede()`, and `join_follow()` functions. These functions are similar
to the overlapping functions, in that they return the query ranges that are
nearest (or precede or follow) the subject ranges and add metadata from
the subject ranges when the query is a nearest neighbour of the subject. Like
the overlap joins, these functions can modified with suffixes to find nearest
neighbours that are left, right, upstream or downstream of the subject.

The pairing operations, `pair_overlap()`, `pair_nearest()`, `pair_follow()`, 
and `pair_precede()` are similar to the join operation but instead of returning
a _Ranges_, they pair up the subject and query _Ranges_ objects into a _DataFrame_,
alongside their metadata columns. This data structure is similar to the _Pairs_
data structure in the _S4Vectors_ @R-S4vectors package or the BED-PE file format.

@@

The _plyranges_ API has an expressive algebra for performing arithmetic on Ranges via
the verbs `set_width()` and `stretch()`. As the names suggest `set_width()`
modifies the width of a Ranges object, while `stretch` extends the start and
end of a Ranges object. These can be chained with the anchoring functions
`anchor_start()`, `anchor_end()`, `anchor_center()`, `anchor_3p()` or 
`anchor_5p()`, which fix the coordinates of a Ranges object in place. 
Moreover, the `shift_` and `flank_` family of functions can be used to shift
all coordinates in a Ranges object or generate flanking regions from a Ranges 
object to the left, right, upstream or downstream of the input. Unlike, the
Bioconductor API, _plyranges_ makes it explicit via function calls whether to
take into account the strand information of a _Ranges_ object.

The methods `as_granges()` and `as_iranges()`
for constructing _Ranges_ from tabular data structures, such as the _data.frame_ in
base R. These methods use non-standard evaluation so columns in a _data.frame_
can be modified before a _Ranges_ object is created. The API also has
convenience methods for annotating or extracting annotations from _GRanges_
objects with the `set_genome_info()` and `get_genome_info()` functions.

@@

@@

We have designed every function to perform a single
action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the `dplyr` package
and implements context specific verbs for transforming _Ranges_ objects (described in
table ). Each verb preserves the semantics of _Ranges_ object and via
duck typing works with derivatives of the _Ranges_ class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a  vocabulary of several verbs, rather 
than having to memorise function names that are nouns. Moreover, this approach 
strongly constrasts the `GenomicRanges/IRanges` OO interface, which emphasises 
the use of setter and getter methods. In that interface, core components and 
metadata are updated via replacement methods (requiring knowledge of the
class components), while our interface requires only a single call to `mutate()`
to perform the modification.

@@

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tibble)
feature_tbl <- tribble(
  ~Verb, ~Action, ~Description,
  "`mutate()`", "modifies columns", "Takes a _Ranges_ object and a set of name-value pairs and generates a new _Ranges_ object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
 "`filter()`", "subset rows", "Takes a _Ranges_ object and a set of logical expressions and restricts the _Ranges_ object to where all logical expressions are true.",
 "`summarise()`", "aggregate one or more columns by a function", "Takes a _Ranges_ object and a set of name-value pairs and aggregates the _Ranges_ according to functions evaluated in the name-value pairs.",
"`group_by()`", "partition a _Ranges_ object by a column", "Create an implicit grouping of _Ranges_ object by one or more columns. This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
"`select()`", "select columns of a _Ranges_ objects", "Drop or reorder metadata columns.",
"`arrange()`", "sort a _Ranges_ object by columns", "Sort a _Ranges_ object by named variables",
"`disjoin_ranges()`", "aggregate one more columns of a _Ranges_ object over the union of end coordinates", "Takes a _Ranges_ object and a set of name-value pairs and expands the _Ranges_ object by taking the union of end coordinates. For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
"`reduce_ranges()`", "aggregate one more columns of a _Ranges_ object by merging overlapping and neighbouring ranges.", "Takes a _Ranges_ object and a set of name-value pairs and reduces the _Ranges_ object by merging overlapping and neighouring ranges. For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
)
```

 (that
are named as nouns) with many arguments (which is required in the existing 
Bioconductor packages)
 This is possible because every 
function call is an endomorphism: when the input is _Ranges_ object the output
will also be a _Ranges_ object. One advantage of endomorphism is that it 
does not  require any additional learning of classes beyond _Ranges_ and 
the _DataFrame_ classes. This strongly deviates from the design of 
the _Ranges_ Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables efficient computing as users
are exposed to low-level features of its API which _plyranges_ abstracts
away. Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure.

The combination of verbs we have defined above cover most operations that
can be performed in the original _IRanges_ and _GenomicRanges_ packages
without the user being exposed to new classes. In those packages to perform
most operations requires users to learn many classes and perform 
additional manipulations to return the results of their computation back to a
_Ranges_ object.

@@

A caveat to constructing a compatible API with `dplyr` is that 
`plyranges` makes extensive use of non-standard evaluation in R
(achieved via the `rlang` package @R-rlang). Simply, this
means that computations are performed and evaluated in the context of the 
_Ranges_ objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with `plyranges` becomes more difficult 
because a user needs to know about non-standard evaluation and user defined
functions can be difficult to debug.

>>>>

---
title: "plyranges: a grammar for transforming genomics data" 
author:
  - name: Stuart Lee
    email: stuart.lee1@monash.edu
    affiliation: Monash University
  - name: Michael Lawrence
    email: michael.lawrence@gene.com
    affiliation: Genentech
  - name: Di Cook
    email: di.cook@monash.edu
    affiliation: Monash University
address:
  - code: Monash University
    address: Department of Econometrics and Business Statistics, Clayton, Victoria, Australia
  - code: Genentech
    address: Bioinformatics and Computational Biology, Genentech, Inc., South San Francisco, California, United States of America 
abstract: |
    The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments. However, there is a cognitive load placed on a user in learning a data abstraction and understanding its appropriate use. Throughout a standard workflow, a user must navigate and know many of these abstractions to perform an genomic analysis task, when a single data abstraction, a GRanges object will suffice. The GRanges class naturally represent genomic intervals and their associated measurements. By recognising that the GRanges class follows 'tidy' data principles we have created a grammar of genomic data transformation. The grammar defines verbs for performing actions on and between genomic interval data. It provides a principled way
    of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows. We have implemented this grammar as an Bioconductor/R package called plyranges.

@@

Another example of our algebra altering object semantics, while maintaining 
cohesion is through the 'group_by' operator. Like anchoring, this operator 
decorates a GRanges object with a column name (or names) that defines a 
partitioning of the  GRanges by the unique values in the column(s). Functions 
defined in the  `plyranges` DSL that perform restriction, aggregation, or 
column modfication still perform those singular tasks, however grouping changes
how those tasks are performed.

Our algebra recasts the actions of finding overlaps or nearest neightbours 
between two genomic regions as a relational join operator. The join operator 
acts on two GRanges object, a query and a subject. The join operator is 
relational in the sense that metadata from the query and subject ranges is 
retained in the joined range.  The commonality between
all join operators in the `plyranges` DSL is to use the index of where in
the query range the subject range 'hits' the query range as a primary key 
(figure *need to make figure*). The notion of 'hits' and hence
the resulting key is determined by the type of join (preceding, nearest, follows
overlaps).

We extend this idea by introducing three types of overlap joins: inner,
intersect and left. The inner join considers a hit to be any overlap between
subject and query and expands the query range to have length equal to the 
number of hits. The intersect join uses the same definition of a hit as
the inner join but returns a range where the start and end coordinates
are the intersect of coordinates in the subject that hit the query. Finally,
the overlap left join is akin to left outer join in Cobb's relational algebra:
it performs an overlap inner join but also returns all query ranges that
are not hit by the subject.

The behaviour of a join operator can be further altered with additional
suffixes by restricting or expanding how a subject 'hits' a query. For example,
for overlap joins we can use suffixes to encapsulate Allan's Interval Algebra:
to consider subject 'hits' that are entirely 'within' a query range, the 'within'
prefix is used. We also use the 'directed' suffix to explicitly include 
notions of strandedness in our algebra.

comparsion to Bioconductor API here?

@@

@@

The methods `as_granges()` and `as_iranges()`
for constructing GRanges from tabular data structures, such as the _data.frame_ in
base R. These methods use non-standard evaluation so columns in a _data.frame_
can be modified before a GRanges object is created. The API also has
convenience methods for annotating or extracting annotations from _GRanges_
objects with the `set_genome_info()` and `get_genome_info()` functions.

@@

 (that
are named as nouns) with many arguments (which is required in the existing 
Bioconductor packages)
 This is possible because every 
function call is an endomorphism: when the input is GRanges object the output
will also be a GRanges object. One advantage of endomorphism is that it 
does not  require any additional learning of classes beyond GRanges and 
the _DataFrame_ classes. This strongly deviates from the design of 
the GRanges Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables efficient computing as users
are exposed to low-level features of its API which _plyranges_ abstracts
away. Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure.

The combination of verbs we have defined above cover most operations that
can be performed in the original _IRanges_ and _GenomicRanges_ packages
without the user being exposed to new classes. In those packages to perform
most operations requires users to learn many classes and perform 
additional manipulations to return the results of their computation back to a
GRanges object.

@@

We have designed every function to perform a single
action on a GRanges object. The `plyranges` DSL implments the core verbs from the `dplyr` package
and implements context specific verbs for transforming GRanges objects (described in
table ). Each verb preserves the semantics of GRanges object and via
duck typing works with derivatives of the GRanges class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a  vocabulary of several verbs, rather 
than having to memorise function names that are nouns. Moreover, this approach 
strongly constrasts the `GenomicRanges/IRanges` OO interface, which emphasises 
the use of setter and getter methods. In that interface, core components and 
metadata are updated via replacement methods (requiring knowledge of the
class components), while our interface requires only a single call to `mutate()`
to perform the modification.

@@

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tibble)
feature_tbl <- tribble(
  ~Verb, ~Action, ~Description,
  "`mutate()`", "modifies columns", "Takes a GRanges object and a set of name-value pairs and generates a new GRanges object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
 "`filter()`", "subset rows", "Takes a GRanges object and a set of logical expressions and restricts the GRanges object to where all logical expressions are true.",
 "`summarise()`", "aggregate one or more columns by a function", "Takes a GRanges object and a set of name-value pairs and aggregates the GRanges according to functions evaluated in the name-value pairs.",
"`group_by()`", "partition a GRanges object by a column", "Create an implicit grouping of GRanges object by one or more columns. This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
"`select()`", "select columns of a GRanges objects", "Drop or reorder metadata columns.",
"`arrange()`", "sort a GRanges object by columns", "Sort a GRanges object by named variables",
"`disjoin_ranges()`", "aggregate one more columns of a GRanges object over the union of end coordinates", "Takes a GRanges object and a set of name-value pairs and expands the GRanges object by taking the union of end coordinates. For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
"`reduce_ranges()`", "aggregate one more columns of a GRanges object by merging overlapping and neighbouring ranges.", "Takes a GRanges object and a set of name-value pairs and reduces the GRanges object by merging overlapping and neighouring ranges. For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
)
```

@@

A caveat to constructing a compatible API with `dplyr` is that 
`plyranges` makes extensive use of non-standard evaluation in R
(achieved via the `rlang` package @R-rlang). Simply, this
means that computations are performed and evaluated in the context of the 
GRanges objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with `plyranges` becomes more difficult 
because a user needs to know about non-standard evaluation and user defined
functions can be difficult to debug.
